To be FAIR: Theory Specification Needs an Update

### 2025-06-20 manuscript.Rmd
Update manuscript.Rmd
extended info on phonological loop

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

>>>>

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
<span id="phonoloop">Take for example the phonological loop in Baddeley's working memory model [@baddeley1992]. The idea behind it is that verbal information is kept in memory via a rehearsal loop with limited capacity. The verbal description of the phonological loop in Baddeley's theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate of information, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].</span>
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.
